### 2018-04-08 Writing/disparity_review.tex
rewritten intro

\noindent{\Large \bf Variations in multidimensional space.} 
% NC: Not sure about this title, especially now we are just doing a mini series. I think it really needs disparity in the title somewhere.
\bigskip

@@

%NC: OK I think this paragraph needs some rewriting. At present it's unclear to me what you're trying to tell me with this paragraph. Are you trying to tell me about the history of disparity? Are you trying to define it and tell me about the debates surrounding that? I think mostly you're aiming for the latter, but this needs to be much clearer. Try and be a bit more upfront about what you're doing, and don't be afraid of short, clear sentences without any fancy words. i.e. epistemiology and ontology helps no-one! Also note that you start by saying disparity = morphology, then you say it could be anything but people argue about it, then you say it can be anything. So perhaps you need to start broad and then say one type of disparity, and the most commonly analysed, is morphological, and that's what most people would think of when hearing the term. But it could apply to anything.

%1 -
% History of disparity methods
%Since the time of the ancient, it has been observed that variation in morphologies often follows discrete patterns rather than a continuous gradient of forms. % NC: Since the time of the ancients is super clunky. I've tried something different
Even classical scholars recognised that variation in the shape and form of organisms often follows discrete patterns rather than a continuous gradient.
% NC: changed the next bit as you need to define disparity/introduce it. As previously written it was hiding in a bracket.
Disparity analysis, the study of differences among these clusters of morphological groups, was popularised in seminal palaeobiology papers in the 1990s [Runnegar 1987, Gould, Foote, Webb, etc.].
Since then, disparity has been used to describe biological, phenotypic, and/or trait variation without strict distinction between the ontology and epistemology (see Semantics Box) 
% NC: Or glossary rather than semantics?
% NC: I have no idea what you mean with this ontology and epistemiology comment. 
Over the last 30 years, disparity has been most commonly used in the context of morphological disparity.
The impressive amount of literature published since then tackles a vast array of biological questions, using different (sometimes contrasted) methods but often consists studies of the phenotype at a macroevolutionary level (i.e. species or above).
To date, there is no consensus on whether disparity should be restricted and specific to these kind of studies or whether it should be used as a broader umbrella term designating studies of multivariate variation, not restricted in its questions, methods and disciplines.
Throughout this review, we aim to highlight the diversity of disparity analyses and how they can help understanding or provide a structure to biodiversity.
In essence, we suggest that disparity is a pattern of variation.

%2§ what use: commonly used
% NC: I feel like this perhaps need to fit in to the paragraph above, where you're saying it can be used for lots of things? 
The description of these patterns % NC: what patterns?
 has been used to tackle a vast array of questions.
For example, it has been used (1) in Evo-devo to tackle questions of
%evo-devo
Developmental constraints [CITE];
Relationship between form and function and evolution functional traits [CITE] or
Scaling micro to macro [CITE];
% Ecology
(2) in ecology or palaeoecology to study:
intra-specific competition [CITE],
niche occupants, speciation and competition[CITE] or
macroecology (disparity in space)[CITE];
% Evolution
(3) on in macroevolution to study:
response to extinction and selectivity [CITE],
intra-specific competitive replacements [CITE] or
more globally, tempo and mode of evolution [CITE].

%3§ problems %NC: Could maybe put the subheading "best practices" here?
Disparity studies have a great deal of promise in helping us to understand the evolution of biological diversity.
With the expansion of methodological approaches in the last decade, however, we also identify three main issues that, if not tackled could reduce the power of such analysis.
First, many recent studies have lost focus; they are no longer hypothesis-driven, instead they are undertaken to characterise biological variation for its own sake.
Second, even where there is a clear question being explored, the data being used for disparity analyses is inappropriate and often driven by the availability of large datasets.
Third, the methods being used to analyse these datasets may not always be appropriate for the question and/or data at hand.
Here we deal with some of these issues and present best practice guidelines to allow researchers to optimise their questions before performing a disparity analysis.

@@

\section{Use appropriate methods} 
In the vast majority of morphological disparity analyses, the methodological approach will be multidimensional.
Multidimensional analysis can be a powerful tool for analysing complex datasets but also comes with many mathematical caveats and limitations.
As with any great power, it comes with great responsibility. % NC: Ha are we citing spiderman too? Awesome
The first step in any statistical analysis should be to visualise the data to understand its distribution and limitations.
Unfortunately, we can only perceive a maximum of three dimensions at the same time.
Therefore one of the main challenges in multidimensional analysis is how to reduce the number of dimensions, both for visualisation and analysis (see box \ref{box_ordination}).

@@

\subsubsection{Models}
One 
What are our null hypotheses?
Joe’s simulations’ stuff - phylogeny can be used to build a null.One basic null is that disparity increases with diversity; how do we diverge from that (show these morpho/genome distance divergence/convergence).
We can have a null model that is based on simple Brownian increase in disparity (gradual). 
Null: is the null increase in disparity dependent on a random walk and on a tree shape?
Maybe we don’t want to think about a null model but rather think about good models to fit. 
We haven’t explored much about fitting models like explaining holes, effects of extinction etc…
Dealing with uncertainty? Bootstrap Bayesian G matrices \& P matrices!

>>>>

\noindent{\Large \bf Disparity in morphological disparity analysis} 
% NC: Not sure about this title, especially now we are just doing a mini series. I think it really needs disparity in the title somewhere.
%TG: back to the drawing board. Maybe something with "best practices" as well.
\bigskip

@@

Even classical scholars recognised that variation in the shape, form and functions of organisms often follows discrete patterns rather than a continuous gradients.
These patterns can be pictured as multidimensional spaces encompassing all the possible life forms where each individual region are the result of a discrete combination of many variables.
For example, the mammalian carnivores design space can be seen as a combination of phylogenetic variables and their associated shared traits (i.e. the synapomorphies of the order Carnivora) as well as a combination of ecological, life history and behavioural variables (i.e. the ecological guild of meat eating mammals).

One commonly used approach to analyse these discrete patterns and to understand the processes that led to it is to analyse the disparity of the morphological charactersitics of the group, often referred to as morphological disparity analysis.
This type of analysis was popularised around the 1990s [Runnegar 1987, Gould, Foote, Webb, etc.] by studying the fossil record and focusing on broad questions such how does diversity evolve, are some body types more common that others or can shape evolve in all ``directions'' (of the multidimensional space).
Since then, an impressive amount of literature and the equally vast amount of methods have been used and the term ``disparity'' has become a close synonym of ``analysis of patterns of morphological diversity'' and has been used in a range of sub-fields in biology.
Disparity analysis has been used to tackle a vast array of questions.
For example, it has been used (1) in Evo-devo to tackle questions of
%evo-devo
Developmental constraints [CITE];
Relationship between form and function and evolution functional traits [CITE] or
Scaling micro to macro [CITE];
% Ecology
(2) in ecology or palaeoecology to study:
intra-specific competition [CITE],
niche occupants, speciation and competition[CITE] or
macroecology (disparity in space)[CITE];
% Evolution
(3) on in macroevolution to study:
response to extinction and selectivity [CITE],
intra-specific competitive replacements [CITE] or
more globally, tempo and mode of evolution [CITE].

To date, however, there is no consensus on whether disparity should be restricted and specific to these kind of studies or whether it should be used as a broader umbrella term designating studies of multivariate variation, not restricted in its questions, methods and disciplines.
Throughout this review, we aim to highlight the diversity of disparity analyses and how they can help understanding or provide a structure to biodiversity.
In essence, we suggest that disparity is a pattern of variation.
This diversity (and disparity) of methods to analysis disparity provides biologist with a powerful and comprehensive tool %TG: comprehensive in the sense "global"/"universal" not sure if that's the right word here.
that has a great deal of promise in helping us to understand the evolution of biological diversity.
However, as with any great power, it comes with great responsibility.

With the expansion of methodological approaches in the last decade, we also identify three main issues that, if not tackled could reduce the power of such analysis.
First, many recent studies have lost focus; they are no longer hypothesis-driven, instead they are undertaken to characterise biological variation for its own sake.
Second, even where there is a clear question being explored, the data being used for disparity analyses is inappropriate and often driven by the availability of large datasets.
Third, the methods being used to analyse these datasets may not always be appropriate for the question and/or data at hand.
Here we deal with some of these issues and present best practice guidelines to allow researchers to optimise their questions before performing a disparity analysis.

@@

\section{Use appropriate methods} 
In the vast majority of morphological disparity analyses, the methodological approach will be multidimensional.
Multidimensional analysis can be a powerful tool for analysing complex datasets but also comes with many mathematical caveats and limitations.
The first step in any statistical analysis should be to visualise the data to understand its distribution and limitations.
Unfortunately, we can only perceive a maximum of three dimensions at the same time.
Therefore one of the main challenges in multidimensional analysis is how to reduce the number of dimensions, both for visualisation and analysis (see box \ref{box_ordination}).

@@

\subsubsection{Models}
What are our null hypotheses?
Joe’s simulations’ stuff - phylogeny can be used to build a null.One basic null is that disparity increases with diversity; how do we diverge from that (show these morpho/genome distance divergence/convergence).
We can have a null model that is based on simple Brownian increase in disparity (gradual). 
Null: is the null increase in disparity dependent on a random walk and on a tree shape?
Maybe we don’t want to think about a null model but rather think about good models to fit. 
We haven’t explored much about fitting models like explaining holes, effects of extinction etc…
Dealing with uncertainty? Bootstrap Bayesian G matrices \& P matrices!
